\stktokens{}: Enforcing Well-bracketed Control Flow and Stack Encapsulation using Linear Capabilities

### 2018-07-06 fully-abstract-comp/popl2019/paper.tex
typo
Updated step relation to be displayed as inference rule
Added call code and short description.

\section{A Capability Machine with Sealing and Linear Capabilities}
\label{sec:cap-mach-w-seal-and-lin}
\begin{figure}[tb]
  \centering
  \[
  \arraycolsep=1.4pt
  \begin{array}{rrcl l rrcl}
   \addrbnf,\basebnf \in & \Addr & \defeq & \nats & \phantom{mak} & \sealbasebnf, \sigma \in & \Seal & \defeq & \nats\\
    \aendbnf \in & \Addr \cup \{\infty \} & & & & \sealendbnf \in & \Seal \cup \{\infty \}& &\\
    \permbnf \in& \Perm & \defbnf & \rwx \mid \rx \mid \rw \mid \ro \mid \noperm & & &\linbnf & \defbnf & \linear \mid \normal\\
    \vsc \in &\SealableCaps&\defbnf & \multicolumn{6}{l}{((\permbnf,\linbnf),\basebnf,\aendbnf,\addrbnf) \mid \seal{\sealbasebnf,\sealendbnf,\sigma}}\\
    c \in&\Caps& \defbnf &  \SealableCaps \mid \sealed{\sigma,\scbnf} & & w \in &\Word & \defeq & \ints \uplus \Caps\\ 
    r \in& \RegName & \defbnf & \multicolumn{6}{l}{\pcreg \mid \rretd \mid \rretc \mid \rstk \mid \rdata \mid \rtmp{1} \mid \rtmp{2} \mid \dots} \\
    \reg \in &\Reg & \defeq & \RegName \fun \Word & & \mem \in&\Mem & \defeq & \Addr \fun \Word\\
    &\MemSeg & \defeq & \Addr \parfun \Word & & \Phi \in & \ExecConf & \defeq & \Mem \times \Reg\\
    &\Conf & \defeq & \multicolumn{6}{l}{\ExecConf \cup \{\failed\} \cup \{\halted\} }
  \end{array}
\]
\[
  \arraycolsep=1.4pt
\begin{array}{rcl}
\multicolumn{3}{l}{    \arraycolsep=0pt
      \com{r} \in  \tRegName \hspace{2.5cm}   \com{\rn} \defbnf \com{r} \mid \nats
}\\
  \Instr &\defbnf & \tjmp{\com{r}} \mid \tjnz{\com{r}}{\com{\rn}} \mid \tmove{\com{r}}{\com{\rn}} \mid \tload{\com{r}}{\com{r}} \mid \tstore{\com{r}}{\com{r}} \mid \tplus{\com{r}}{\com{\rn}}{\com{\rn}} \mid \tminus{\com{r}}{\com{\rn}}{\com{\rn}} \mid\\
         & & \tlt{\com{r}}{\com{\rn}}{\com{\rn}} \mid \tisptr{\com{r}}{\com{r}} \mid\tgetp{\com{r}}{\com{r}} \mid \tgetlin{\com{r}}{\com{r}} \mid \tgetb{\com{r}}{\com{r}} \mid \tgete{\com{r}}{\com{r}} \mid \tgeta{\com{r}}{\com{r}}  \mid \\
  & & \tcca{\com{r}}{\com{n\rn}} \mid \tsetatob{\com{r}} \mid \trestrict{\com{r}}{\com{\rn}} \mid \tcseal{\com{r}}{\com{r}} \mid \txjmp{\com{r}}{\com{r}} \mid  \tsplit{\com{r}}{\com{r}}{\com{r}}{\com{\rn}} \mid\\ 
      & & \tsplice{\com{r}}{\com{r}}{\com{r}} \mid \tfail \mid \thalt 
\end{array}
\]
  \caption{The syntax of our capbility machine with seals and linera capbilities.}
  \label{fig:target-syntax}
\end{figure}
% Brief intro to section and this cap machine:
% Cap machine inspired by CHERI, but with linear capabilities
In this section, we introduce a simple but representative capability machine with linear capabilities.
We will later introduce a capability machine similar to the one in this section, so to easily distinguish the two, we will call this one \trgcm{} (Linear Capability Machine).
\trgcm{} is mainly inspired by CHERI~\citep{watson_cheri:_2015} with linear capabilities as the main exception.
In order to avoid details uninteresting to the focus of this work, \trgcm{} assumes an infinite address and unbounded integers.
\lau{Do we have a good reason to assume an infinite address space for this work?
  We do not model malloc, so we do not need to worry about what happens when malloc runs out of memory.}

@@

@@

    \Phi \step
    \sem{\decInstr{\Phi.\mem(\aaddr)}}(\Phi) &
    \arraycolsep=0pt
      \begin{array}[t]{l}
        \text{if }\Phi(\pcreg) = ((\perm,\_),\baddr,\eaddr,\aaddr) \wedge
        \baddr \le \aaddr \le \eaddr \wedge \perm \in \{\rwx,\rx\}
      \end{array} \\

@@

% Summary of CC
In the this section, we have described our proposed calling convention, \stktokens{}.
In this paragraph we summarize it:
\begin{description}
\item[Check the base address of the stack capability before and after calls] 
\item[Make sure that local stack frames are non-empty]
\item[Create token and data return capability on call] split the stack capability in two to get a stack capability for you local stack frame and a stack capability for the unused part of the stack. The stack capability for your local stack frame is sealed and used for the data part of the return pair.
\item[Create code return capability on call] Seal the old program pointer.
\item[Reasonable use of seals] return seals are only used to seal old program pointers, all return seals are only used for one call site, and return seals are not leaked.
\end{description}

@@

>>>>

\section{A Capability Machine with Sealing and Linear Capabilities}
\label{sec:cap-mach-w-seal-and-lin}
\begin{figure}[tb]
  \centering
  \[
  \arraycolsep=1.4pt
  \begin{array}{rrcl l rrcl}
   \addrbnf,\basebnf \in & \Addr & \defeq & \nats & \phantom{mak} & \sealbasebnf, \sigma \in & \Seal & \defeq & \nats\\
    \aendbnf \in & \Addr \cup \{\infty \} & & & & \sealendbnf \in & \Seal \cup \{\infty \}& &\\
    \permbnf \in& \Perm & \defbnf & \rwx \mid \rx \mid \rw \mid \ro \mid \noperm & & &\linbnf & \defbnf & \linear \mid \normal\\
    \vsc \in &\SealableCaps&\defbnf & \multicolumn{6}{l}{((\permbnf,\linbnf),\basebnf,\aendbnf,\addrbnf) \mid \seal{\sealbasebnf,\sealendbnf,\sigma}}\\
    c \in&\Caps& \defbnf &  \SealableCaps \mid \sealed{\sigma,\scbnf} & & w \in &\Word & \defeq & \ints \uplus \Caps\\ 
    r \in& \RegName & \defbnf & \multicolumn{6}{l}{\pcreg \mid \rretd \mid \rretc \mid \rstk \mid \rdata \mid \rtmp{1} \mid \rtmp{2} \mid \dots} \\
    \reg \in &\Reg & \defeq & \RegName \fun \Word & & \mem \in&\Mem & \defeq & \Addr \fun \Word\\
    &\MemSeg & \defeq & \Addr \parfun \Word & & \Phi \in & \ExecConf & \defeq & \Mem \times \Reg\\
    &\Conf & \defeq & \multicolumn{6}{l}{\ExecConf \cup \{\failed\} \cup \{\halted\} }
  \end{array}
\]
\[
  \arraycolsep=1.4pt
\begin{array}{rcl}
\multicolumn{3}{l}{    \arraycolsep=0pt
      \com{r} \in  \tRegName \hspace{2.5cm}   \com{\rn} \defbnf \com{r} \mid \nats
}\\
  \Instr &\defbnf & \tjmp{\com{r}} \mid \tjnz{\com{r}}{\com{\rn}} \mid \tmove{\com{r}}{\com{\rn}} \mid \tload{\com{r}}{\com{r}} \mid \tstore{\com{r}}{\com{r}} \mid \tplus{\com{r}}{\com{\rn}}{\com{\rn}} \mid \tminus{\com{r}}{\com{\rn}}{\com{\rn}} \mid\\
         & & \tlt{\com{r}}{\com{\rn}}{\com{\rn}} \mid \tisptr{\com{r}}{\com{r}} \mid\tgetp{\com{r}}{\com{r}} \mid \tgetlin{\com{r}}{\com{r}} \mid \tgetb{\com{r}}{\com{r}} \mid \tgete{\com{r}}{\com{r}} \mid \tgeta{\com{r}}{\com{r}}  \mid \\
  & & \tcca{\com{r}}{\com{n\rn}} \mid \tsetatob{\com{r}} \mid \trestrict{\com{r}}{\com{\rn}} \mid \tcseal{\com{r}}{\com{r}} \mid \txjmp{\com{r}}{\com{r}} \mid  \tsplit{\com{r}}{\com{r}}{\com{r}}{\com{\rn}} \mid\\ 
      & & \tsplice{\com{r}}{\com{r}}{\com{r}} \mid \tfail \mid \thalt 
\end{array}
\]
  \caption{The syntax of our capbility machine with seals and linear capbilities.}
  \label{fig:target-syntax}
\end{figure}
% Brief intro to section and this cap machine:
% Cap machine inspired by CHERI, but with linear capabilities
In this section, we introduce a simple but representative capability machine with linear capabilities.
We will later introduce a capability machine similar to the one in this section, so to easily distinguish the two, we will call this one \trgcm{} (Linear Capability Machine).
\trgcm{} is mainly inspired by CHERI~\citep{watson_cheri:_2015} with linear capabilities as the main exception.
In order to avoid details uninteresting to the focus of this work, \trgcm{} assumes an infinite address and unbounded integers.
\lau{Do we have a good reason to assume an infinite address space for this work?
  We do not model malloc, so we do not need to worry about what happens when malloc runs out of memory.}

@@

  \begin{mathpar}
    \inference{\Phi(\pcreg) = ((\perm,\_),\baddr,\eaddr,\aaddr) \\
        \baddr \le \aaddr \le \eaddr \and \perm \in \{\rwx,\rx\}
 }{ \Phi \step \sem{\decInstr{\Phi.\mem(\aaddr)}}(\Phi) }
              \and
              \inference{ \neg (\Phi \step \Phi') \and \Phi' \neq \failed}
                        {\Phi \step \failed }
  \end{mathpar}

@@

  %   \Phi \step
  % \begin{cases}
  %   \sem{\decInstr{\Phi.\mem(\aaddr)}}(\Phi) &
  %   \arraycolsep=0pt
  %     \begin{array}[t]{l}
  %       \text{if }\Phi(\pcreg) = ((\perm,\_),\baddr,\eaddr,\aaddr) \wedge
  %       \baddr \le \aaddr \le \eaddr \wedge \perm \in \{\rwx,\rx\}
  %     \end{array} \\
  %     \failed & \totherwise
  % \end{cases}\\

@@

% Summary of CC
In the this section, we have described our proposed calling convention, \stktokens{}. The measures taken are:
\begin{enumerate}
\item \label{item:check-stkb} Check the base address of the stack capability before and after calls
\item \label{item:non-empty-sf} Make sure that local stack frames are non-empty
\item \label{item:return-data} Create token and data return capability on call split the stack capability in two to get a stack capability for you local stack frame and a stack capability for the unused part of the stack. The stack capability for your local stack frame is sealed and used for the data part of the return pair.
\item \label{item:return-code} Create code return capability on call Seal the old program pointer.
\item Reasonable use of seals return seals are only used to seal old program pointers, all return seals are only used for one call site, and return seals are not leaked.
\end{enumerate}
Item~\ref{item:check-stkb}-\ref{item:return-code} of \stktokens{} are captured by the code in Figure~\ref{fig:call-code} with the exception of checking stack base before calls. We do not include this, because it only needs to happen once between two calls which means that the check after a call suffices if the base of the stack pointer is not changed subsequently.

@@

\begin{figure}[htb]
  \centering
{\footnotesize
\[
  \begin{array}{r >{$}p{3.5cm}<{$} l >{$}p{3.5cm}<{$} l l}
    1: & \tmove{\rtmp{1}}{42} & 10: &\tcseal{\rretd}{\rtmp{1}} & 19: &\tcca{\rtmp{2}}{5} \\
    2: &\tstore{\rstk}{\rtmp{1}} & 11: &\tmove{\rretc}{\pcreg} & 20: &\tjnz{\rtmp{2}}{\rtmp{1}}\\
    3: &\tcca{\rstk}{(-1)} & 12: & \tcca{\rretc}{5} & 21: &\tcca{\rtmp{2}}{1} \\
    4: &\tgeta{\rtmp{1}}{\rstk} & 13: & \tcseal{\rretc}{\rtmp{1}} & 22: &   \tjmp{\rtmp{2}} \\
    5: &\tsplit{\rstk}{\rretd}{\rstk}{\rtmp{1}} & 14: & \tmove{\rtmp{1}}{0} & 23: &    \tfail \\
    6: &\tmove{\rtmp{1}}{\pcreg} & 15: & \txjmp{r_1}{r_2} & 24: &   \tsplice{\rstk}{\rstk}{\rdata} \\
    7: &\tcca{\rtmp{1}}{(\offpc - 5)} & 16: & \tgetb{\rtmp{1}}{\rstk} & 25: &    \tcca{\rstk}{1}\\
    8: &\tload{\rtmp{1}}{\rtmp{1}} & 17: &\tminus{\rtmp{1}}{\rtmp{1}}{\stkb} & 26: &    \tmove{\rtmp{2}}{0} \\
    9: &\tcca{\rtmp{1}}{\offsigma} & 18: &\tmove{\rtmp{2}}{\pcreg} & &
  \end{array}
\]
}
  \caption{The instructions that constitutes a call. The variable $\offpc$ is the offset from the first instruction of the call to the set of seals it uses. The variable $\offsigma$ is the offset in the set of seals. The constant $\stkb$ is the globally agreed on stack base. There are a number of ``magic'' numbers in the code: line 1: $42$, garbage data to ensure a non-empty stack. Line 7: $-5$, offset to the first instruction (from the point the capability was moved out of the $\pcreg$-register). Line 12: $5$, offset to the return address. Line 19: $5$, offset to fail. Line 21: offset to address after fail.}
  \label{fig:call-code}
\end{figure}
